A dynamic programming algorithm to assess the reliability of inferences from network enrichment analysis

### 2018-05-02 doc/genesetdp.tex
Spelling

\documentclass[a4paper,american]{lipics-v2016}
%\usepackage[USenglish]{babel}
%\usepackage{microtype}
\bibliographystyle{plainurl}
\title{A dynamic programing algorithm to assess the reliability of inferences from network enrichment analysis}
\titlerunning{Dynamic Programing for Network Enrichment inferences} %optional, in case that the title is too long; the running title should fit into the top page column

@@

Gene enrichment analysis (GEA) is commonly used to infer function from sets of analytes such as genes, transcripts proteins or metabolites\cite{tavazoie1999systematic,khatri2012ten}. The technique estimates which functional modules, such as complexes or pathways, that are overrepresentated among a set of identified analytes. One prominent application of the technique is expression analysis, where GEA is regularly used to assess alternation in pathway activity by examining significantly different concentrations of analytes between biological conditions, such as disease state or treatment group.

@@

A limiting factor of GEA are the pathway definition databases. While \url{http://pathguide.org} list 708 databases of
pathway definitions\cite{bader2006pathguide},  some chritique has been voiced concerning the completeness and rigor of current pathway databases. Hence efforts have been directed to designing methods that extend the pathway definitions, by functional association networks, like STRING\cite{szklarczyk2014string}. Instead of directly examining the overlap between the query and pathway sets, one can evaluate the number of links in the functional association network that connect the query and pathway set\cite{shojaie2010network, alexeyenko2012network, glaab2012enrichnet, mccormack2013statistical, ogris2016novel, signorelli2016neat}. This is known as Network Enrichment Analysis (NEA).

@@

Here, we present a dynamic programing algorithm, that we dubbed GeneSetDP, that calculate the full score distribution of any query of a given size. The algorithm enables us to calculate well defined and exact statistics of inferences from NEA.  Using simulations we demonstrated that the algorithm produced unbiased statistics.

@@

We also downloaded the NEA example given at the BinoX web site. The example files include a pathway definition file that groups $6819$ genes into $289$ human pathways, and a network definition file that, after tresholding with a score of $0.7$, yeilds $1244992$ links bewtween those genes.

As an representative example for we chose the Glycolysis/Gluconeogenesis pathway, as it contained a number of genes that coincided with the median number of genes of the pathway definition file.

@@

We implemented a Python program that read network and pathway definition files, and scored query sets according to Equation \ref{eq:sum_i}, the Dynamic Programming algorithm described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}. We downloaded pathway and network definitions from the BinoX's website and used the same network treshold as BinoX default ($0.7$).

@@

In order to test the statistical calibration of our method, we calculated $p$~values for 10000 random picks of gene sets from the investigated genome. As our selection is random, and following our null hypothesis, we expected the resulting $p$~values to be uniformly distributed. To test this we plotted the $p$~values against their quantile in Figure \ref{fig:calibration}. As a comparison we also added a calibration curve for the popular NEA method BinoX\cite{ogris2016novel}.

@@

%\includegraphics{img/FIXME.png}
  \caption{{\bf Calibration of GeneSetDP.} We selected 10000 random sets from the investigated genome definition, and calculated their $p$ values with GeneSetDP against the Glycolysis/Gluconeogenesis pathway. To test the $p$~values uniformity we plotted them against their Normalized rank. We also added the calibration curve of BinoX, evaluated on the same random sets.}
  \label{fig:calibration}
\end{figure}

@@

It should be noted that our outlined method a score distribution could work well with Monte Carlo-sampling as well, by sampling random gene sets of size $Q$, which would be faster and easier to understand than the theory of how to permute networks.

Here we made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$ : ``The Q query genes are randomly selected from the genes in the genome''. Previous implementation of NEA in the literature otherwise use Monte Carlo-simulation to determine parameters for various types parametric distributions, by perturbations of the investigated network. In practice such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use to calculate accurate statistics. In general, it is easier to understand null models that relates to user actions, than null models that relates to model parameters.

>>>>

\documentclass[a4paper,american]{lipics-v2016}
%\usepackage[USenglish]{babel}
%\usepackage{microtype}
\bibliographystyle{plainurl}
\title{A dynamic programming algorithm to assess the reliability of inferences from network enrichment analysis}
\titlerunning{Dynamic Programming for Inferences from Network Enrichment Analysis} %optional, in case that the title is too long; the running title should fit into the top page column

@@

Gene enrichment analysis (GEA) is commonly used to infer function from sets of analytes such as genes, transcripts proteins or metabolites\cite{tavazoie1999systematic,khatri2012ten}. The technique estimates which functional modules, such as complexes or pathways, that are overrepresented among a set of identified analytes. One prominent application of the technique is expression analysis, where GEA is regularly used to assess alternation in pathway activity by examining significantly different concentrations of analytes between biological conditions, such as disease state or treatment group.

@@

A limiting factor of GEA is the pathway definition databases. While \url{http://pathguide.org} list 708 databases of pathway definitions\cite{bader2006pathguide},  some critique has been voiced concerning the completeness and rigor of current pathway databases. Hence efforts have been directed to designing methods that extend the pathway definitions, by functional association networks, like STRING\cite{szklarczyk2014string}. Instead of directly examining the overlap between the query and pathway sets, one can evaluate the number of links in the functional association network that connect the query and pathway set\cite{shojaie2010network, alexeyenko2012network, glaab2012enrichnet, mccormack2013statistical, ogris2016novel, signorelli2016neat}. This is known as Network Enrichment Analysis (NEA).

@@

Here, we present a dynamic programming algorithm, that we dubbed GeneSetDP, that calculates the score distribution of any query of a given size. The algorithm enables us to calculate well defined statistics of inferences from NEA.  Using simulations we demonstrated that the algorithm produced unbiased statistics.

@@

We also downloaded the NEA example given at the BinoX web site. The example files include a pathway definition file that groups $6819$ genes into $289$ human pathways, and a network definition file that, after thresholding with a score of $0.7$, yeilds $1244992$ links between those genes.

As a representative example for we chose the Glycolysis/Gluconeogenesis pathway, as it contained a number of genes that coincided with the median number of genes of the pathway definition file.

@@

We implemented a Python program that read network and pathway definition files and scored query sets according to Equation \ref{eq:sum_i}, the Dynamic Programming algorithm described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}. We downloaded pathway and network definitions from the BinoX's website and used the same network treshold as BinoX default ($0.7$).

@@

In order to test the statistical calibration of our method, we calculated $p$~values for 10000 random picks of gene sets from the investigated genome. As our selection is random and following our null hypothesis, we expected the resulting $p$~values to be uniformly distributed. To test this we plotted the $p$~values against their quantile in Figure \ref{fig:calibration}. As a comparison, we also added a calibration curve for the popular NEA method BinoX\cite{ogris2016novel}.

@@

%\includegraphics{img/FIXME.png}
  \caption{{\bf Calibration of GeneSetDP.} We selected 10000 random sets from the investigated genome definition and calculated their $p$ values with GeneSetDP against the Glycolysis/Gluconeogenesis pathway. To test the $p$~values uniformity we plotted them against their Normalized rank. We also added the calibration curve of BinoX, evaluated on the same random sets.}
  \label{fig:calibration}
\end{figure}

@@

It should be noted that our outlined method a score distribution could work well with Monte Carlo-sampling as well, by sampling random gene sets of size $Q$, which would be faster and easier to understand then the theory of how to permute networks.

Here we made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$ : ``The Q query genes are randomly selected from the genes in the genome''. Previous implementations of NEA in the literature otherwise use Monte Carlo-simulation to determine parameters for various types parametric distributions, by perturbations of the investigated network. In practice, such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use to calculate accurate statistics. In general, it is easier to understand null models that relate to user actions, than null models that relate to model parameters.
